The pursuit of general-purpose robotics requires agents capable of executing high-level natural language instructions in unstructured environments. Recent advancements in Model-Based Reinforcement Learning (MBRL) have demonstrated that planning within compressed latent manifolds—using techniques like latent Model Predictive Control (MPC)—yields highly sample-efficient and robust continuous control. However, a fundamental limitation restricts the deployment of these optimization-based frameworks for language-guided tasks: standard latent MPC algorithms require an explicit representation of the target goal state at inference time to formulate the cost function. When provided only with a linguistic instruction, this dense visual goal representation is absent, breaking the optimization loop.

To bridge this gap, this thesis proposes a task-conditioned, energy-based latent predictive model that acts as an "imagination engine" for downstream control. By fusing high-dimensional visual observations via a Vision Transformer (\texttt{ViT}) and semantic task descriptions via a Parameter-Efficient Fine-Tuned (PEFT) CLIP encoder, the model autoregressively generates reference latent trajectories conditioned purely on language. A critical challenge in Joint-Embedding Predictive Architectures (JEPAs) is the risk of latent collapse during sequential rollouts. We address this by integrating Sketched Isotropic Gaussian Regularization (SIGReg) alongside a multi-step rollout objective, mathematically ensuring that the latent manifold maintains its information-theoretic entropy and isotropic distribution.

Beyond control, establishing trust in autonomous robotic systems requires transparency in their internal decision-making processes. Because latent MPC "thinks" in uninterpretable, high-dimensional spaces, evaluating the safety of an intended action trajectory is inherently difficult. To address this, we train a convolutional decoder to project the model's autoregressive rollouts back into pixel space, effectively allowing human operators to "read the robot's thoughts" and verify its intent prior to physical execution. We evaluate the framework on a custom robotic manipulation dataset. Quantitative analysis utilizing a novel Task Sensitivity Index demonstrates the model's robust discriminative capability in grounding semantics. Qualitative geometric analyses via PCA, alongside decoded visual imaginations, reveal that the model synthesizes divergent, causally-accurate branches for counterfactual instructions. Ultimately, this work establishes a transparent representational foundation that provides the necessary predictive signals for future language-guided, trustworthy latent MPC frameworks.

