\chapter{Introduction}
\label{sec:introduction}

\section{Motivation}

The quest for Autonomous Machine Intelligence (AMI) in robotics is largely driven by the desire to build agents capable of understanding high-level human instructions and executing them in complex, unstructured environments~\cite{lecun2022path}. Traditionally, reinforcement learning (RL) has tackled robotic control through model-free paradigms, mapping high-dimensional visual observations directly to motor actions~\cite{nair2018visual}. However, these approaches are notoriously sample-inefficient and struggle with generalization, as they merely memorize reactive behaviors rather than understanding the underlying physics of the environment.

To address these limitations, the field has increasingly pivoted toward Model-Based Reinforcement Learning (MBRL) and the development of ``world models''~\cite{haarnoja2017reinforcement}. World models learn the transition dynamics of an environment, allowing an agent to simulate or ``imagine'' the consequences of its actions before executing them. Because predicting future states in the raw pixel space is computationally prohibitive and prone to capturing task-irrelevant noise, state-of-the-art world models operate within compressed latent spaces. Frameworks such as TD-MPC~\cite{Hansen2022tdmpc}, its successor TD-MPC2~\cite{hansen2024tdmpc2}, and recent Joint-Embedding Predictive Architectures (JEPAs) like LeWorldModel~\cite{Maes2026LeWorldModelSE} and DINO-WM~\cite{zhou2025dinowm} have demonstrated that planning in a latent manifold can yield highly sample-efficient and robust continuous control.

By leveraging Model Predictive Control (MPC) on top of these latent world models, agents can optimize action trajectories at runtime. This optimization-based approach offers two distinct advantages over reactive policies: first, it allows the system to allocate more computational resources to difficult maneuvers and less to trivial ones; second, by continuously re-planning in a closed loop, the agent gains strong dynamic adaptation capabilities, correcting for environmental stochasticity and unexpected disturbances on the fly.

\section{Problem Statement}

Despite the success of latent MPC frameworks, a fundamental limitation restricts their deployment in general-purpose, language-guided robotics. Current goal-conditioned predictive control methods fundamentally require a latent representation of the target goal at inference time to formulate the optimization cost function. In practical deployment scenarios, however, a user simply provides a high-level linguistic instruction (e.g., ``Pick up the red cube''); the agent is not provided with an explicit visual image or latent vector of the future success state.

Without a goal representation, the MPC optimizer lacks a reference trajectory or terminal state to guide the action sequence. Consequently, there is a critical missing link between abstract natural language commands and the dense, step-by-step latent tracking signals required by closed-loop optimal control algorithms.

To bridge this gap, this thesis aims to develop a task-conditioned, energy-based latent predictive model. Instead of relying on a pre-defined future goal image, the proposed model must understand the task-conditioned dynamics of the environment derived solely from the current visual observation and a linguistic prompt. By autoregressively unrolling a sequence of future latent states conditioned on the task, the model acts as an ``imagination engine.'' This rolled-out latent trajectory provides the exact step-by-step reference signal needed for a downstream latent MPC optimizer to track.

Furthermore, beyond optimal control, the deployment of autonomous robots in human-centric environments demands transparency and trust. Because latent MBRL operates entirely within a high-dimensional, abstract vector space, the robot's internal planning process acts as an uninterpretable ``black box''~\cite{milani2024explainable}. Human operators have no mechanism to verify what the robot intends to do before it physically executes the action. For a robotic system to be truly trustworthy, it should not only provide signals for the control algorithm but also offer human-interpretable explanations of its ``thoughts'' and intended goals~\cite{tabrez2019explanation, ha2018worldmodels}.

Finally, predicting sequential futures in a joint-embedding architecture introduces a severe risk of \textit{latent collapse}—a failure mode where the encoder maps all observations to a constant vector to trivially minimize prediction error. While techniques to solve latent collapse exist in self-supervised learning, they must be carefully integrated into the autoregressive predictive framework without destroying the temporal coherence of the rolled-out trajectories. Therefore, the core problem is to build a language-conditioned latent trajectory generator that maintains a non-degenerate, information-rich manifold suitable for both robust downstream control and human-verifiable intent.

\section{Contributions}

This thesis addresses the aforementioned challenges by formulating and evaluating a generative framework for task-conditioned latent predictive modeling. By treating the transition dynamics as an Energy-Based Model (EBM), we provide a principled approach to evaluating the compatibility of state trajectories with language instructions. The primary contributions of this work are as follows:

\begin{enumerate}
	\item \textbf{Formulation of a Task-Conditioned Latent EBM:} We design a predictive architecture that fuses high-dimensional visual observations (via Vision Transformers) and semantic task descriptions (via PEFT-adapted CLIP encoders) into a shared latent manifold. This model serves as a forward dynamics engine capable of autoregressively generating reference trajectories without requiring visual goal states at inference time.

	\item \textbf{Integration of Information-Theoretic Regularization for Sequential Rollouts:} To make the generated latent trajectories viable for downstream MPC tracking, we integrate Sketched Isotropic Gaussian Regularization (SIGReg)~\cite{lejepa} with a multi-step rollout loss. We demonstrate that this combination provably mitigates latent collapse while enforcing long-horizon physical coherence.

	\item \textbf{Visualizing Latent Intent for Trustworthy Robotics:} To break the "black-box" nature of latent planning, we train a convolutional decoder that maps the latent manifold back to pixel space. This allows human operators to visually decode the autoregressive rollouts, providing a human-interpretable window into the robot's ``imagination.'' This mechanism serves as a foundation for transparent, explainable robotic control where intent can be verified prior to physical execution.

	\item \textbf{Empirical Validation of Task Sensitivity:} We provide quantitative and qualitative analyses demonstrating the model's discriminative grounding of language. Through the formulation of a Task Sensitivity Index and PCA-based manifold visualizations, we show that the model successfully synthesizes divergent, causally-correct latent trajectories based purely on differing counterfactual linguistic instructions.
\end{enumerate}

While the direct integration of this model with a real-time MPC optimizer is reserved for future work, the framework developed herein provides the necessary representational foundation and predictive signals required for language-guided, closed-loop optimal control.

\section*{Structure of the Thesis}

The remainder of this thesis is structured as follows. \cref{sec:background} provides a comprehensive background and literature review, covering advancements in goal-conditioned reinforcement learning, world models, and non-contrastive self-supervised representation learning. \cref{sec:methodology} details the methodology, including the mathematical formulation of the energy-based model, the autoregressive rollout mechanism, and the regularization techniques employed. \cref{sec:experiments} outlines the experimental setup, detailing the dataset, training configurations, and architectural baselines. \cref{sec:results} presents the results and discussion, evaluating the model's task sensitivity, human-interpretable decodings, latent manifold geometry, and the efficacy of the collapse-mitigation strategies. Finally, \cref{sec:conclusion} concludes the thesis and outlines avenues for future research, particularly focusing on the integration of this framework into closed-loop control systems.
